Field and experimental data together with computational

### 2022-07-20 manuscript_PLOS/manuscript_PLOS.tex
New version of the manuscript

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Field and experimental data together with computational
models reveal how cleaner fish adjust decisions in a biological
market} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Andrés E. Quiñones\textsuperscript{\getAff{a}}\textsuperscript{*},
Zegni Triki\textsuperscript{\getAff{b}},
Redouan Bshary\textsuperscript{\getAff{Institute of Biology, University
of Neuchatel}}\\
\bigskip
\textbf{\getAff{a}}Institute of Biology,University of Neuchatel ,Rue
Emile-Argand 11, 2000-CH Neuchatel, Switzerland\\
\textbf{\getAff{b}}Department of Zoology, Stockholm University, Svante
Arrheniusväg 18 B, 106 91 Stockholm, Sweden,\\
\bigskip
* Corresponding author: andresqp@gmail.com\\
\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
While it is generally straightforward to quantify individual performance
in cognitive experiments, identifying the underlying cognitive processes
remains a major challenge. Often, different mechanistic underpinnings
yield similar performances, and Lloyd Morgan's cannon warrants
acceptance of the simpler explanation. Alternatively, when the different
mechanisms interact with environmental conditions, variation in
performance across environments might allow to statistically infer the
mechanism responsible. We illustrate this point by fitting computational
models to experimental data on performance by wild-caught cleaner fish
\emph{Labroides dimidiatus} in an ephemeral reward task, as well as
cleaner and client fish densities from the locations of capture. Using
Bayesian statistics to fit the model parameters to performance data
revealed that cleaner fish most likely estimate future consequences of
an action, while it appears unlikely that the removal of the ephemeral
reward acts as psychological punishment (negative reinforcement).
Incorporating future consequences also yields performances that can be
considered the result of locally optimal decision-rules, in contrast to
the negative reinforcement mechanism. We argue that the combination of
computational models with data is a powerful tool to infer the
mechanistic underpinnings of cognitive performance.

@@

When researchers aimed at testing wild-caught cleaner fish, as well as
individuals from other species, abilities to prioritize visitor over
resident clients in a lab-based paradigm, they used Plexiglas plates of
different colours and shapes offering the same amount of food as
surrogates for visitor and resident clients. One plate is made to behave
like a resident, i.e.~it remains until the cleaner fish has eaten off
it. The other plate was made to behave like a visitor, i.e.~it is
removed if not inspected first. This laboratory paradigm originally
inspired by the cleaning market of cleaner fish is more generally
referred as the ephemeral reward task. Individuals are assessed in their
capacity to develop a preference for the ephemeral option, which gives
them access to two reward units. Cleaners learned to prefer the visitor
plate and hence obtained the double amount of food {[}12{]}.
Furthermore, adult cleaner fish outperformed various primates as well as
rats and pigeons in this original version of the ephemeral reward task
{[}13,14{]}; yet the African grey parrots solve this task as well
{[}15{]}. This between variation in performance, and primates' known
cognitive abilities {[}\textbf{reader\_Evolution\_2001?}{]}, shows that
performance in the task cannot be readily explained by classifying
species as having high and low cognitive abilities. Instead, this
between species variation suggests that solving the ephemeral reward
task requires a set of cognitive mechanisms that are not present in all
the tested taxa, or were not used under the experimental set-up.

Recent research on the cognitive abilities needed to solve the ephemeral
reward task has used a broad approach as proposed by Shettleworth
{[}16{]}, who defined cognition as including all ways in which animals
acquire information through the senses, process, retain and decide to
act on it. Indeed, perception of relevant cues is of major importance
for individual performance. For example, some animals find relevant
information in the plates {[}17{]} while others find it in the food
{[}18,19{]}. However, identifying cues as salient is not the only
challenge of the ephemeral reward task, as revealed by proximate
learning models. Such models allow varying the cognitive tool kit and
evaluating which minimal kit is necessary to solve the task at hand
(e.g. {[}20{]}). Applied to the ephemeral reward task, learning models
showed that basic reinforcement learning does not suffice to solve the
ephemeral reward task {[}21,22{]}. This is particularly so when models
assume the more complex natural situation in which cleaner fish face not
only resident-visitor pairs but also visitor-visitor and
resident-resident pairs, as well a resident alone or a visitor alone. To
be able to give visitors priority over residents, cleaners need to be
able to assess a client's value separately for the three possible
scenarios (alone, paired with a fish with the same strategic option,
paired with a fish with the alternative strategic option) {[}22{]}. The
ability to distinguish and value one stimulus differently alone from
compound versions of it has been termed configurational learning,
chunking, or segmentation (see references in {[}21{]}).

In addition to configurational learning, cleaners also need to account
for the future consequences of current decisions. In the model by
Quiñones \emph{et al.} {[}22{]}, this could be achieved in two
non-mutually exclusive ways: through low temporal discounting of future
effects, also termed `chaining' {[}23{]}; and/or through perceiving a
visitor client leaving as psychological punishment (i.e.~as a negative
reinforcer). Low temporal discounting is when individuals include in
their valuation of an action the reward effects that this will have in
the future. This is done by combining in a single valuation the reward
obtained in the current time with all the reward that comes after,
discounting for how far in the future reward is accrued. `Chaining' the
reward of these different time steps allows individuals to take actions
that increase the long-term reward at the sacrifice of short term
considerations {[}23{]}. Even though, `chaining' can be readily
implemented computationally in learning models {[}23,24{]}, cognitively
it seems to be a complex adaptation {[}2{]}. On the other hand, using
client behaviour as a negative reinforcer is, in principle, easier to
implement. Negative reinforcement is part of the ubiquitous learning
mechanism termed operant conditioning {[}25,26{]}. Thus, the standard
logic of Lloyd Morgan's cannon demands that operant conditioning as the
simpler explanation is to be accepted by default. Ideally, however, the
two mechanisms should be evaluated in light of how well they explain the
available data. Note that different fields interested in cognition and
decision making use different words to refer to negative reinforcers
{[}22,24{]}. Here, for the sake of simplicity and clarity, we will use
the word `penalty' to refer to this mechanism which includes a negative
reinforcer.

Over the last decade, our team tested over a hundred wild-caught cleaner
fish in the exact same paradigm of the ephemeral reward task
{[}13,27--30{]}. These fish often come from different reef locations
around the study field site at Lizard Island, Great Barrier Reef in
Australia. Further investigation of the local eco-sociological
conditions revealed that cleaner and client fish population densities
have a substantial impact on cleaner fish performance in the task.
Cleaner fish from reef sites with relatively low densities were more
likely to fail at solving the task {[}27--29{]}. The explanation for
this pattern is twofold. First, low cleaner fish population densities
imply low supply and high demand for cleaning services. Under such
conditions, there are fewer occasions under which failing to give
priority to a visitor translates to an empty cleaning station. Second
and also related to the state of the cleaning market, visitor clients at
low cleaner fish population densities are less likely to exert their
partner choice options and are hence more likely to wait for inspection
if not given cleaning priority {[}30{]}. In the absence of partner
choice cleaner fish should not give priority to visitors any more.

@@

The model consists of a set of individual-based simulations where
individuals, representing cleaner fish, face a series of choices between
two options, which simulate the natural conditions of the cleaning
market. Individuals experience a series of discrete time points in which
they face different `states,' defined by the number and category of
client fish (visitor or residents) inviting for cleaning services. There
are six possible states: zero clients, one resident, one visitor,
resident-resident, visitor-visitor, and resident-visitor. The
probability of each state is largely determined by the relative
abundance of cleaner fish, residents and visitors, but to some degree by
cleaner fish choices when it faces the resident-visitor combination.
This is because residents are willing to queue for cleaning service;
while visitors leave the queue (with a certain probability) when made to
wait. Individuals obtain a fixed reward from cleaning a client fish
regardless of the category. Every time individuals face and make a
choice they update the probability of making that same choice. The
update is based on the difference between the expected reward and the
obtained reward (prediction error) {[}24,31{]}. Given the new
information, the update is carried in the direction that leads to more
reward being obtained, given the new information. In the long run, the
probability of choosing a visitor over a resident converges in the
model. To which probability the model will converge depends on the
relative abundance of cleaners, visitors and residents; as well as on
the probability of visitors leaving the cleaning station when made to
wait. Further details of the model implementation can be found in
Quiñones \emph{et al} {[}22{]}.

@@

The empirical data were collected between 2010 and 2019 always during
the austral winter months June to August from a total of five study reef
sites (Corner Beach-CB, Horseshoe-HS, Mermaid Cove-MC, Northern
Horseshoe-NHS, and The Crest-TC) at Lizard Island
(\(14.6682^\circ S, 145.4604^\circ E\)), Great Barrier Reef, Australia.
The data consist of three sets: fish censuses, field observations of
cleaner-client interactions to quantify the probability of visitors
leaving if made to wait, and the performance of wild-caught cleaner fish
in the ephemeral reward test. In total, we have twelve site/year data
sets for fish censuses and corresponding performance in the lab test.
Thus, some sites were sampled more than once. To estimate the population
density of cleaner fish and their clients at a given site in a given
year, Triki \emph{et al.} {[}29{]} used a series of ten transects of
\(30m\) each. Observers swam along the transect lines placed on the reef
and first counted the visible large-bodied adult fish (species with
total length \(TL \geq 10cm\)) including cleaner fish on a width of
\(5m\), and then on the return individuals of small-bodied fish species
(\(TL < 10 cm\))\\
on a width of \(1 m\) (see Triki et al.~2018 for further details on fish
censuses data collection). We then scaled the counts of cleaner fish,
small-bodied, and large-bodied clients fish densities per \(100 m^2\).
From the study by Triki \emph{et al} {[}29{]} cleaner fish and
large-bodied client populations densities were highly correlated, and
only the former was hence used in the analyses as representative of both
measures.

The field observation data consisted of video recordings/encodings of
the cleaner-client cleaning interactions. There were videos from eight
cleaners per site/year of a duration of \(30 min\) each. Triki \emph{et
al} extracted information from every event wherein a visitor client was
made to wait in favour of another client (visitor or resident), and
noted whether or not the visitor left or queued for the cleaning service
{[}29,30{]}.

The cognitive performance data was from a total of 120 cleaners (10
individuals per 12 site/year) tested in the ephemeral reward task
{[}29,30{]}. Authors housed all captured cleaners individually in glass
aquaria ( \(62cm \times 27cm \times 37 cm\) ) and provided them with PVC
pipes (\(10 cm \times 1 cm\)) as shelters. The task consisted of
exposing the cleaner fish to substitute models of client fish in the
form of two \emph{Plexiglas} plates offering the same amount of food
(one item of mashed prawn). The two plates differed in colour and
pattern (horizontal green stripes or vertical pink stripes) but had
equal size (\(10 cm \times 7 cm\)). Importantly, the two plates played
different roles as either a visitor (ephemeral food source) or resident
(permanent food source). That is, if a cleaner fish inspected the
resident plate first, the experimenter withdrew the visitor plate out of
the aquarium as a consequence. Choosing first the visitor plate,
however, granted access to both plates. The equal size of the plates
forced cleaner fish to learn to give service priority to the visitor
plate based solely on the behaviour-cue of the plates rather than
size-cue {[}17{]}. Triki \emph{et al.} {[}29,30{]} tested the fish for a
maximum of 200 trials with 20 trials a day, 10 trials in the morning and
10 trials in the afternoon. They randomized and counterbalanced the
plates' spatial location (i.e.~left or right) between trials. Similarly,
they counterbalanced the plates' decoration (colour and pattern) and the
plates' role (visitor or resident) between the tested fish. In the
original studies, once a fish reach a learning criterion, that is,
performing significantly above chance level (\(> 50%
\), \(p-value ≥ 0.05\)), they passed to a reversal version of the task
where the roles of the visitor/resident Plexiglas plates were swapped
(see Triki et al.~2019, 2020). Here, we used instead a subset of these
data in order to have an idea of cleaner fish preferences for the
visitor plate, even if they do not reach the learning criterion within
200 trials. To do so, we first extracted the trial-by-trial outcomes
from the last two sessions (20 trials) of those who never reached the
learning criterion for visitor plate (\(N = 45\) cleaner fish). For
those who reached the learning criterion at some point during the test
and passed to a reversal phase, we extracted the trial-by-trial outcomes
from the last session (10 trials) before passing to reversal and the
last 10 trials they were exposed to in the test (\(N = 75\) cleaner
fish). We chose a combination of initial and reversal to quantify
preference for the visitor client, given that it matches well the
criteria chosen in previous analysis of the ephemeral reward task
experimental set-up {[}29,30{]}. In the supplementary material (Fig.
\ref{fig:rawdata}) we show how our choice maps to the previously used
criteria. For comparison, we also show a quantification of visitor
preference based only on the initial phase of the task.

@@

The aim of the analyses is to fit the key model parameters \(\gamma\)
and \(\eta\), to the empirical data from Triki \emph{et al.} {[}29,30{]}
to test whether each or a combination of these effects is a better
explanation for the pattern seen in the data. We used the ecological
variables: cleaners, visitor clients, resident clients abundances and
visitor clients leaving probability,\\
as input to the models. As the response variable, we used cleaners'
preference for visitors over residents in the ephemeral reward task.
Finally, we used the preference for visitors resulting from the model
simulations as the prediction for the response variable. We kept all
other parameter values used for the model simulations constant, see
Table \ref{tab:param}.

@@

Once we calculated the relative abundances, we obtained predictions from
the model for each one of the locations and ran the Markov Chain. We
started the chain with random values for the parameters of interest,
then ran the computational model once for every reef site using as input
the ecological explanatory variables. The model outputs the probability
of choosing the visitor for each location \(p_i\), where \(i\) is the
index for the 12 locations. Assuming a binomial distribution, the
probability that each of the cleaner fish 20 choices in the ephemeral
reward task was generated by the model is given by
\(\binom{n}{k_ij}p^{k_{ij}}_i (1-p_i)^{n-k_{ij}}\); where \(k_{ij}\) is
the number of times that cleaner fish \(j\) in reef site \(i\) chose the
visitor over the resident; and \(n=20\) (due to our choice of using 20
choices per cleaner fish). By taking the natural logarithm and summing
over all the cleaner fish and reef sites, we obtained the log-likelihood
of the data given the model and parameters. We then proposed a new set
of parameters drawn from a uniform distribution centred around the old
parameter set. The amplitude of the uniform distribution used for each
parameter can be found in Table \ref{tab:param}. Subsequently, we ran
the model and calculated the likelihood with the new parameter set. We
then used the ratio of the two likelihoods to choose which parameter set
to keep. New parameter sets with a higher likelihood than the old set
replaced old ones, and those with a lower likelihood replaced current
ones with a probability equal to the log-likelihood ratio. Given that we
only used the likelihoods in the decision, we used an uninformative
prior. Once we decided whether the new parameter set would replace the
old one, we ran the model again to sample the likelihood distribution of
the parameter set. We then started the cycle again by proposing a new
set of parameters and repeated the process for \(1e^5\) steps. We ran 5
independent chains, discarded the first \(1000\) samples of each chain
as burn in, and after that, we kept 1 in every 100 samples to avoid
autocorrelation. The collection of parameters sets kept in all chains
approximates the posterior distribution. We coded the model as well as
the fitting algorithm in \emph{c++}, the diagnostics and visualization
in R {[}32{]}. All codes are accessible at
\url{https://github.com/andreseqp/learnDatamodel}

To compare the fit of the three alternative models, we used the
distribution of pseudo \(R^2\) proposed by Mcfadden {[}33{]}. Mcfadden's
pseudo \(R^2\) is a standard measure of fit for logistic regression. In
that context, \(pseudo-R^2\) uses the log-likelihood of the data given
the model, relative to the log-likelihood of the data given a model
without covariates, as a measure of fit. Our model is not a logistic
regression, therefore we measured the \(pseudo-R^2\) in relation to the
log-likelihood of a model with parameters \(\gamma\) and \(\eta\) set to
zero. This, in practice, amounts to a model that has a neutral
preference between the two options. We computed the pseudo \(R^2\) for
all the samples of the posterior from the MCMC. Thus, we used these
distributions of pseudo \(R^2\)'s as a measure of fit.

@@

While all three models captured well the positive relationship between
visitor leaving behaviour and cleaner fish performance in the market
task {[}29{]}, only the chaining mechanism predicted that cleaner fish
performance in the task should be low in habitats with low
cleaner-to-client ratios, regardless of the visitor leaving probability.
In contrast, models including negative reward predicted the highest
performance in the ephemeral reward task when relative cleaner fish
abundance is low, particularly together with a high probability of
visitor leaving (Fig. \ref{fig:pred}). Low relative cleaner fish
abundances mean the market has an excess of demand for cleaning
services. In the models, this translates to a cleaning station that is
frequently full. Thus, when a visitor leaves, it is likely that the
cleaner fish will have access to another client in the next step.
Therefore, there will not be much difference in future reward between
choosing a visitor and a resident, and cleaners will not develop a
preference for the visitor in these conditions. On the other hand, the
effect of negative reward on cleaner fish preference is the opposite, as
in a busy cleaning station, to that of chaining. cleaner fish will get
more often the resident-visitor state and will develop a preference for
the visitor faster. At high cleaner fish abundances, the
resident-visitor state becomes so rare that neither mechanism is very
efficient at generating a preference for visitors. When facing the
resident-visitor choice, it is still best to choose the visitor;
however, the learning machinery will not be able to develop this
preference efficiently. Overall, the models suggest that chaining is the
cognitive mechanism that allows cleaner fish to adaptively adjust to
their biological market ecological conditions.

>>>>

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Field and experimental data together with computational
models reveal how cleaner fish adjust decisions in a biological
market} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Andrés E. Quiñones\textsuperscript{\getAff{a}}\textsuperscript{*},
Zegni Triki\textsuperscript{\getAff{b}},
Redouan Bshary\textsuperscript{\getAff{a}}\\
\bigskip
\textbf{\getAff{a}}Institute of Biology, University of Neuchatel,
Neuchatel, Switzerland\\
\textbf{\getAff{b}}Department of Zoology, Stockholm University,
Stockholm, Sweden\\
\bigskip
* Corresponding author: andresqp@gmail.com\\
\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
While it is generally straightforward to quantify individual performance
in cognitive experiments, identifying the underlying cognitive processes
remains a major challenge. Often, different mechanistic underpinnings
yield similar performances, and Lloyd Morgan's cannon warrants
acceptance of the simpler explanation. Alternatively, when the different
mechanisms interact with environmental conditions, variation in
performance across environments might allow to statistically infer the
mechanism responsible. We illustrate this point by fitting computational
models to experimental data on performance by wild-caught cleaner fish
\emph{Labroides dimidiatus} in an ephemeral reward task, as well as
cleaner and client fish densities from the locations of capture. Using
Bayesian statistics to fit the model parameters to performance data
revealed that cleaner fish most likely estimate future consequences of
an action, while it appears unlikely that the removal of the ephemeral
reward acts as psychological punishment (negative reinforcement).
Incorporating future consequences also yields performances that can be
considered the result of locally optimal decision-rules, in contrast to
the negative reinforcement mechanism. We argue that the combination of
computational models with data is a powerful tool to infer the
mechanistic underpinnings of cognitive performance.

@@

When researchers aimed at testing wild-caught cleaner fish, as well as
individuals from other species, abilities to prioritize visitor over
resident clients in a lab-based paradigm, they used Plexiglas plates of
different colours and shapes offering the same amount of food as
surrogates for visitor and resident clients. One plate is made to behave
like a resident, i.e.~it remains until the cleaner fish has eaten off
it. The other plate was made to behave like a visitor, i.e.~it is
removed if not inspected first. This laboratory paradigm originally
inspired by the cleaning market of cleaner fish is more generally
referred as the ephemeral reward task. Individuals are assessed in their
capacity to develop a preference for the ephemeral option, which gives
them access to two reward units. Cleaners learned to prefer the visitor
plate and hence obtained the double amount of food {[}12{]}.
Furthermore, adult cleaner fish outperformed various primates as well as
rats and pigeons in this original version of the ephemeral reward task
{[}13,14{]}; yet the African grey parrots solve this task as well
{[}15{]}. This between variation in performance, and primates' known
cognitive abilities {[}16{]}, shows that performance in the task cannot
be readily explained by classifying species as having high and low
cognitive abilities. Instead, this between species variation suggests
that solving the ephemeral reward task requires a set of cognitive
mechanisms that are not present in all the tested taxa, or were not used
under the experimental set-up.

Recent research on the cognitive abilities needed to solve the ephemeral
reward task has used a broad approach as proposed by Shettleworth
{[}17{]}, who defined cognition as including all ways in which animals
acquire information through the senses, process, retain and decide to
act on it. Indeed, perception of relevant cues is of major importance
for individual performance. For example, some animals find relevant
information in the plates {[}18{]} while others find it in the food
{[}19,20{]}. However, identifying cues as salient is not the only
challenge of the ephemeral reward task, as revealed by proximate
learning models. Such models allow varying the cognitive tool kit and
evaluating which minimal kit is necessary to solve the task at hand
(e.g. {[}21{]}). Applied to the ephemeral reward task, learning models
showed that basic reinforcement learning does not suffice to solve the
ephemeral reward task {[}22,23{]}. This is particularly so when models
assume the more complex natural situation in which cleaner fish face not
only resident-visitor pairs but also visitor-visitor and
resident-resident pairs, as well a resident alone or a visitor alone. To
be able to give visitors priority over residents, cleaners need to be
able to assess a client's value separately for the three possible
scenarios (alone, paired with a fish with the same strategic option,
paired with a fish with the alternative strategic option) {[}23{]}. The
ability to distinguish and value one stimulus differently alone from
compound versions of it has been termed configurational learning,
chunking, or segmentation (see references in {[}22{]}).

In addition to configurational learning, cleaners also need to account
for the future consequences of current decisions. In the model by
Quiñones \emph{et al.} {[}23{]}, this could be achieved in two
non-mutually exclusive ways: through low temporal discounting of future
effects, also termed `chaining' {[}24{]}; and/or through perceiving a
visitor client leaving as psychological punishment (i.e.~as a negative
reinforcer). Low temporal discounting is when individuals include in
their valuation of an action the reward effects that this will have in
the future. This is done by combining in a single valuation the reward
obtained in the current time with all the reward that comes after,
discounting for how far in the future reward is accrued. `Chaining' the
reward of these different time steps allows individuals to take actions
that increase the long-term reward at the sacrifice of short term
considerations {[}24{]}. Even though, `chaining' can be readily
implemented computationally in learning models {[}24,25{]}, cognitively
it seems to be a complex adaptation {[}2{]}. On the other hand, using
client behaviour as a negative reinforcer is, in principle, easier to
implement. Negative reinforcement is part of the ubiquitous learning
mechanism termed operant conditioning {[}26,27{]}. Thus, the standard
logic of Lloyd Morgan's cannon demands that operant conditioning as the
simpler explanation is to be accepted by default. Ideally, however, the
two mechanisms should be evaluated in light of how well they explain the
available data. Note that different fields interested in cognition and
decision making use different words to refer to negative reinforcers
{[}23,25{]}. Here, for the sake of simplicity and clarity, we will use
the word `penalty' to refer to this mechanism which includes a negative
reinforcer.

Over the last decade, our team tested over a hundred wild-caught cleaner
fish in the exact same paradigm of the ephemeral reward task
{[}13,28--31{]}. These fish often come from different reef locations
around the study field site at Lizard Island, Great Barrier Reef in
Australia. Further investigation of the local eco-sociological
conditions revealed that cleaner and client fish population densities
have a substantial impact on cleaner fish performance in the task.
Cleaner fish from reef sites with relatively low densities were more
likely to fail at solving the task {[}28--30{]}. The explanation for
this pattern is twofold. First, low cleaner fish population densities
imply low supply and high demand for cleaning services. Under such
conditions, there are fewer occasions under which failing to give
priority to a visitor translates to an empty cleaning station. Second
and also related to the state of the cleaning market, visitor clients at
low cleaner fish population densities are less likely to exert their
partner choice options and are hence more likely to wait for inspection
if not given cleaning priority {[}31{]}. In the absence of partner
choice cleaner fish should not give priority to visitors any more.

@@

The model consists of a set of individual-based simulations where
individuals, representing cleaner fish, face a series of choices between
two options, which simulate the natural conditions of the cleaning
market. Individuals experience a series of discrete time points in which
they face different `states,' defined by the number and category of
client fish (visitor or residents) inviting for cleaning services. There
are six possible states: zero clients, one resident, one visitor,
resident-resident, visitor-visitor, and resident-visitor. The
probability of each state is largely determined by the relative
abundance of cleaner fish, residents and visitors, but to some degree by
cleaner fish choices when it faces the resident-visitor combination.
This is because residents are willing to queue for cleaning service;
while visitors leave the queue (with a certain probability) when made to
wait. Individuals obtain a fixed reward from cleaning a client fish
regardless of the category. Every time individuals face and make a
choice they update the probability of making that same choice. The
update is based on the difference between the expected reward and the
obtained reward (prediction error) {[}25,32{]}. Given the new
information, the update is carried in the direction that leads to more
reward being obtained, given the new information. In the long run, the
probability of choosing a visitor over a resident converges in the
model. To which probability the model will converge depends on the
relative abundance of cleaners, visitors and residents; as well as on
the probability of visitors leaving the cleaning station when made to
wait. Further details of the model implementation can be found in
Quiñones \emph{et al} {[}23{]}.

@@

The empirical data were collected between 2010 and 2019 always during
the austral winter months June to August from a total of five study reef
sites (Corner Beach-CB, Horseshoe-HS, Mermaid Cove-MC, Northern
Horseshoe-NHS, and The Crest-TC) at Lizard Island
(\(14.6682^\circ S, 145.4604^\circ E\)), Great Barrier Reef, Australia.
The data consist of three sets: fish censuses, field observations of
cleaner-client interactions to quantify the probability of visitors
leaving if made to wait, and the performance of wild-caught cleaner fish
in the ephemeral reward test. In total, we have twelve site/year data
sets for fish censuses and corresponding performance in the lab test.
Thus, some sites were sampled more than once. To estimate the population
density of cleaner fish and their clients at a given site in a given
year, Triki \emph{et al.} {[}30{]} used a series of ten transects of
\(30m\) each. Observers swam along the transect lines placed on the reef
and first counted the visible large-bodied adult fish (species with
total length \(TL \geq 10cm\)) including cleaner fish on a width of
\(5m\), and then on the return individuals of small-bodied fish species
(\(TL < 10 cm\)) on a width of \(1 m\) (see Triki et al.~2018 for
further details on fish censuses data collection). We then scaled the
counts of cleaner fish, small-bodied, and large-bodied clients fish
densities per \(100 m^2\). From the study by Triki \emph{et al} {[}30{]}
cleaner fish and large-bodied client populations densities were highly
correlated, and only the former was hence used in the analyses as
representative of both measures.

The field observation data consisted of video recordings/encodings of
the cleaner-client cleaning interactions. There were videos from eight
cleaners per site/year of a duration of \(30 min\) each. Triki \emph{et
al} extracted information from every event wherein a visitor client was
made to wait in favour of another client (visitor or resident), and
noted whether or not the visitor left or queued for the cleaning service
{[}30,31{]}.

The cognitive performance data was from a total of 120 cleaners (10
individuals per 12 site/year) tested in the ephemeral reward task
{[}30,31{]}. Authors housed all captured cleaners individually in glass
aquaria ( \(62cm \times 27cm \times 37 cm\) ) and provided them with PVC
pipes (\(10 cm \times 1 cm\)) as shelters. The task consisted of
exposing the cleaner fish to substitute models of client fish in the
form of two \emph{Plexiglas} plates offering the same amount of food
(one item of mashed prawn). The two plates differed in colour and
pattern (horizontal green stripes or vertical pink stripes) but had
equal size (\(10 cm \times 7 cm\)). Importantly, the two plates played
different roles as either a visitor (ephemeral food source) or resident
(permanent food source). That is, if a cleaner fish inspected the
resident plate first, the experimenter withdrew the visitor plate out of
the aquarium as a consequence. Choosing first the visitor plate,
however, granted access to both plates. The equal size of the plates
forced cleaner fish to learn to give service priority to the visitor
plate based solely on the behaviour-cue of the plates rather than
size-cue {[}18{]}. Triki \emph{et al.} {[}30,31{]} tested the fish for a
maximum of 200 trials with 20 trials a day, 10 trials in the morning and
10 trials in the afternoon. They randomized and counterbalanced the
plates' spatial location (i.e.~left or right) between trials. Similarly,
they counterbalanced the plates' decoration (colour and pattern) and the
plates' role (visitor or resident) between the tested fish. In the
original studies, once a fish reach a learning criterion, that is,
performing significantly above chance level (\(> 50%
\), \(p-value ≥ 0.05\)), they passed to a reversal version of the task
where the roles of the visitor/resident Plexiglas plates were swapped
(see Triki et al.~2019, 2020). Here, we used instead a subset of these
data in order to have an idea of cleaner fish preferences for the
visitor plate, even if they do not reach the learning criterion within
200 trials. To do so, we first extracted the trial-by-trial outcomes
from the last two sessions (20 trials) of those who never reached the
learning criterion for visitor plate (\(N = 45\) cleaner fish). For
those who reached the learning criterion at some point during the test
and passed to a reversal phase, we extracted the trial-by-trial outcomes
from the last session (10 trials) before passing to reversal and the
last 10 trials they were exposed to in the test (\(N = 75\) cleaner
fish). We chose a combination of initial and reversal to quantify
preference for the visitor client, given that it matches well the
criteria chosen in previous analysis of the ephemeral reward task
experimental set-up {[}30,31{]}. In the supplementary material (Fig.
\ref{fig:rawdata}) we show how our choice maps to the previously used
criteria. For comparison, we also show a quantification of visitor
preference based only on the initial phase of the task.

\hypertarget{ethics-statement}{%
\subsubsection{Ethics statement}\label{ethics-statement}}

The Animal Ethics Committee of the Queensland government (DAFF) approved
the project (CA 2016/05/970 and CA 2017/05/1063)

@@

The aim of the analyses is to fit the key model parameters \(\gamma\)
and \(\eta\), to the empirical data from Triki \emph{et al.} {[}30,31{]}
to test whether each or a combination of these effects is a better
explanation for the pattern seen in the data. We used the ecological
variables: cleaners, visitor clients, resident clients abundances and
visitor clients leaving probability,\\
as input to the models. As the response variable, we used cleaners'
preference for visitors over residents in the ephemeral reward task.
Finally, we used the preference for visitors resulting from the model
simulations as the prediction for the response variable. We kept all
other parameter values used for the model simulations constant, see
Table \ref{tab:param}.

@@

Once we calculated the relative abundances, we obtained predictions from
the model for each one of the locations and ran the Markov Chain. We
started the chain with random values for the parameters of interest,
then ran the computational model once for every reef site using as input
the ecological explanatory variables. The model outputs the probability
of choosing the visitor for each location \(p_i\), where \(i\) is the
index for the 12 locations. Assuming a binomial distribution, the
probability that each of the cleaner fish 20 choices in the ephemeral
reward task was generated by the model is given by
\(\binom{n}{k_ij}p^{k_{ij}}_i (1-p_i)^{n-k_{ij}}\); where \(k_{ij}\) is
the number of times that cleaner fish \(j\) in reef site \(i\) chose the
visitor over the resident; and \(n=20\) (due to our choice of using 20
choices per cleaner fish). By taking the natural logarithm and summing
over all the cleaner fish and reef sites, we obtained the log-likelihood
of the data given the model and parameters. We then proposed a new set
of parameters drawn from a uniform distribution centred around the old
parameter set. The amplitude of the uniform distribution used for each
parameter can be found in Table \ref{tab:param}. Subsequently, we ran
the model and calculated the likelihood with the new parameter set. We
then used the ratio of the two likelihoods to choose which parameter set
to keep. New parameter sets with a higher likelihood than the old set
replaced old ones, and those with a lower likelihood replaced current
ones with a probability equal to the log-likelihood ratio. Given that we
only used the likelihoods in the decision, we used an uninformative
prior. Once we decided whether the new parameter set would replace the
old one, we ran the model again to sample the likelihood distribution of
the parameter set. We then started the cycle again by proposing a new
set of parameters and repeated the process for \(1e^5\) steps. We ran 5
independent chains, discarded the first \(1000\) samples of each chain
as burn in, and after that, we kept 1 in every 100 samples to avoid
autocorrelation. The collection of parameters sets kept in all chains
approximates the posterior distribution. We coded the model as well as
the fitting algorithm in \emph{c++}, the diagnostics and visualization
in R {[}33{]}. All codes are accessible at \textless DOI:
10.5281/zenodo.6338073\textgreater{}

To compare the fit of the three alternative models, we used the
distribution of pseudo \(R^2\) proposed by Mcfadden {[}34{]}. Mcfadden's
pseudo \(R^2\) is a standard measure of fit for logistic regression. In
that context, \(pseudo-R^2\) uses the log-likelihood of the data given
the model, relative to the log-likelihood of the data given a model
without covariates, as a measure of fit. Our model is not a logistic
regression, therefore we measured the \(pseudo-R^2\) in relation to the
log-likelihood of a model with parameters \(\gamma\) and \(\eta\) set to
zero. This, in practice, amounts to a model that has a neutral
preference between the two options. We computed the pseudo \(R^2\) for
all the samples of the posterior from the MCMC. Thus, we used these
distributions of pseudo \(R^2\)'s as a measure of fit.

@@

While all three models captured well the positive relationship between
visitor leaving behaviour and cleaner fish performance in the market
task {[}30{]}, only the chaining mechanism predicted that cleaner fish
performance in the task should be low in habitats with low
cleaner-to-client ratios, regardless of the visitor leaving probability.
In contrast, models including negative reward predicted the highest
performance in the ephemeral reward task when relative cleaner fish
abundance is low, particularly together with a high probability of
visitor leaving (Fig. \ref{fig:pred}). Low relative cleaner fish
abundances mean the market has an excess of demand for cleaning
services. In the models, this translates to a cleaning station that is
frequently full. Thus, when a visitor leaves, it is likely that the
cleaner fish will have access to another client in the next step.
Therefore, there will not be much difference in future reward between
choosing a visitor and a resident, and cleaners will not develop a
preference for the visitor in these conditions. On the other hand, the
effect of negative reward on cleaner fish preference is the opposite, as
in a busy cleaning station, to that of chaining. cleaner fish will get
more often the resident-visitor state and will develop a preference for
the visitor faster. At high cleaner fish abundances, the
resident-visitor state becomes so rare that neither mechanism is very
efficient at generating a preference for visitors. When facing the
resident-visitor choice, it is still best to choose the visitor;
however, the learning machinery will not be able to develop this
preference efficiently. Overall, the models suggest that chaining is the
cognitive mechanism that allows cleaner fish to adaptively adjust to
their biological market ecological conditions.
